Forecasting hourly hospital admission

### 2021-10-07 paper/paper.Rmd
Minor paper edit
intro added

---
title: Short-term hourly forecasting for care services
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Author2
    email: email2@example.com
    affiliation: University2
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

Demand forecasting is a critical input to inform staffing and scheduling to meet the needs of patients in hospital services. An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives. 
Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow. Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A&E without seen by staff, unexpected return visits to A&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.

An accurate forecast of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of healthcare services provided.

Furthermore, because the consequences of imperfact staffing are asymmetric probabilistic forecasts are necessary to make economic decicions. This asymmetry arrises because it is preferable to incure a small opportunity cost associated unerutilised staff rather than lower service levels if staff levels are insufficient.

Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.

The combination of an hourly forecast demand, incidents being attended, resource availability and delays at hospitals, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.

Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the schediling.

what literature says about hourly forecasting in healthcare?

There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily  arrivals. 
There are limited studies that use
different time series forecasting techniques to predict annual arrival,[24] monthly arrival,[14][25]
daily arrival, [26] and arrival at intervals of 4 to 12 h; [27] however, existing studies have shown
complications in predicting hourly patient ED arrivals[28] and the application of predicting hourly
patient arrivals is not well established. 
Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data. In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services

limitations in current literature?
- Typically focus on daily data, hiding information about peak arrival rates. Hourly data has additional challenge of multiple seasonality.
- Deterministic, are there any examples of probabilistic?

@@

1. We develop a novel methodology to forecast hourly time series for healthcare services combining a distribution fit to each hour of day separately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, ... iv) special events such as public holidays, festive days and rugby and v) weather data.
2. We provide probabilistic forecasts ...
3. We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using ....
4. We provide data and code enabling reproduction and refinement of the proposed approach

# Research background: Hourly forecasting in care services{#lit}

@@

```{r summarylit}
li <- readxl::read_xlsx("../table/table.xlsx")
knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position")) %>% column_spec(5,"15em")
```

@@

The literature review reveals some limitations in forecasting for urgent and emergency care which will be summarised as follows:

- first, 
- second,
- third,

@@

@@

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',fig.height=9, fig.width= 6}

@@

```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
```

@@

To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
\begin{equation}
     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}}
\end{equation}
where $M$ is the metric's value for the method being considered, $M_\text{ref}$ is the metric's value for a reference method, and $M_\text{perf}$ is the metrics value for   the `perfect' method, which is zero in the case of RMSE and Pinball. We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e. positive or negative skill) are significantly different from zero [@Efron1981Bootstrap]. Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.

@@

```{r fig-quantilebias}
## Should reduce the number of models shown here... it's really hard to tell them appart!
knitr::include_graphics(rep("../results/QuantileBias.png"))
```

@@

```{r tab-results}
load("Results_Table.Rda")
knitr::kable(Res_sum[order(Res_sum$`Quantile Bias`)],
  caption = 'Forecast evluation for all methods on test data in decreasing order by Quantile Bias',
  digits=c(NA,3,2,2)
)
```

```{r fig-skill}
knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
```

>>>>

---
title:  Forecasting hourly hospital admission
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Author2
    email: email2@example.com
    affiliation: University2
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients. An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives [@rostami2020anticipating].

Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding]. Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.

There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting]. daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] admission. In this paper, we only focus on forecasting hourly hospital admission. In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services.  Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential. The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[]. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission. Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Aditionally, datasets used in some of these studies are relatively small, short length (e.g. time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications  covered in this paper fails reproducibility principles (Boylan, 2016).

In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:

1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv)

2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;

3. We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e. Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);

4. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.

# Research background: hourly hospital admission forecasting {#lit}

@@

@@

```{r summarylit}
li <- readxl::read_xlsx("../table/table.xlsx")
knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") 
```

@@

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method nor is forecast accuracy reported.

@@

```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width= 7,fig.height=6}
```

@@

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "80%"}

@@

@@

To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using

\begin{equation}
     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} (\#eq:skillscore)
\end{equation}

where $M$ is the metric's value for the method being considered, $M_\text{ref}$ is the metric's value for a reference method, and $M_\text{perf}$ is the metrics value for   the `perfect' method, which is zero in the case of RMSE and Pinball. We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e. positive or negative skill) are significantly different from zero [@Efron1981Bootstrap]. Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.

@@

```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
knitr::include_graphics(rep("../results/QuantileBias.png"))
```

```{r tab-results, out.width= "50%"}
results_table <- read_rds("results_table.rds")
knitr::kable(results_table, booktabs = T, 
             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
```

@@

```{r tab-results-qbias, out.width= "50%"}
results_table_qb <- read_rds("results_table.rds") %>% select(`Quantile Bias`)

knitr::kable(results_table_qb, booktabs = T, 
             linesep = "",caption = "Forecast evluation for all methods on test data in decreasing order by Quantile Bias") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
```

```{r fig-skill, fig.cap= "Skill score ...", fig.align='center',out.width= "90%"}
knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
```
